NumPy---Array Programming in Python

### 2020-01-29 body.tex
Minor edits

For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
SciPy is used to provide tools used in general image processing such as 
filtering and image alignment, while scikit-image---a popular image processing
library that extends SciPy---provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11}---a popular package for complex
network analysis---is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.
% N.B. - most of this comes from introspecting the Image class defined in
% ehtim/image.py and grepping the dependencies in requirements.txt

%Tools for analyzing data from gravitational wave observatories are
%provided by the \code{pycbc} package, which was used in the first
%detection of gravitational waves \cite{abbott2016observation} and in 
%on-going analysis of data from LIGO and Virgo.
%\code{pycbc} also make extensive use of the scientific Python ecosystem.
%The time-series data from the interferometers are stored in NumPy arrays,
%while the \code{scipy.signal} subpackage is used to construct and apply filters to the
%data.
%Matplotlib is used to visualize data at all points along the analysis chain
%including raw data from the instrumentation and the time-frequency 
%visualization of the ``chirps'' that result from mergers of compact binaries
%such as black holes\cite{abbott2016observation} and
%neutron stars\cite{abbott2017multi}.

Similar examples can be found in physics, earth science, biology,
and more.
Researchers in these fields depend on NumPy being correct, reliable,
and repeatable.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
At times, this conflicts with the need to innovate;
resolving it requires balancing trade-offs.
% our focus is ...
% NumPy purpose engages with this problem by ...
% [this also defines the scope of NumPy]
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data in continues
to evolve and change.
% this leads to a proliferation of arrays

@@

Recent years have seen a massive explosion in data science, machine learning,
and artificial intelligence.  NumPy and its API, underlying almost every tool
in the computational toolchain, has become ubiquitous.  Still, NumPy itself
could never hope to satisfy every single specialized need of the community;
due to its in-memory data model, it is limited in its ability to work with very
large datasets, datasets split across multiple systems, and computations that
require graphics processing units (GPUs), tensor processing units (TPUs) or
other specialized hardware for performance reasons.  Pursuing solutions to
these challenges is out of scope for NumPy development.

The community's efforts to fill the resultant gaps therefore led to a
proliferation of array implementations. Each deep learning framework created
its own arrays; PyTorch, Tensorflow, Apache MXNet and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:, 1]} vs.
\code{x.loc[:, 'time']}---by Xarray.

Importantly, some projects chose to follow the NumPy API exactly---with Dask,
CuPy, Sparse and JAX the most prominent examples---while others have
NumPy-inspired semantics but chose to deviate in places.

% \footnote{\url{https://cupy.chainer.org}}
% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
% \footnote{\url{https://mxnet.incubator.apache.org}}
% \footnote{\url{https://dask.org}}
% JAX has a citation, see https://github.com/google/jax#citing-jax
% Xarray has a citation, see http://xarray.pydata.org/en/stable/faq.html#how-should-i-cite-xarray 
% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}

% proliferation of arrays including distributed arrays---which are stored on
% multiple computers, GPU arrays---which utilize graphics processing units for
% fast computation, computation graphs---which postpone and combine calculations
% before executing them efficiently, and sparse arrays---which typically contain
% few non-zero values, and store only those in memory.

%This includes popular tensor computation libraries such as
%CuPy, JAX, and Apache MXNet.
%PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.

@@

% Ralf: FIXME I don't think this is actually important - should work exactly the same
% for users whether a NumPy-compatible API is a subclass or a duck array
Primarily, there
exist two types of Python array objects: (a) arrays that mimic NumPy arrays but are
fundamentally different and (b) subclasses of NumPy arrays.

@@

To facilitate \emph{interoperability}, the NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.

>>>>

For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
SciPy is used to provide tools used in general image processing such as 
filtering and image alignment, while scikit-image---a popular image processing
library that extends SciPy---provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11}---a popular package for complex
network analysis---is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.

Similar examples can be found in physics, earth science, biology,
and more.
Researchers in these fields depend on NumPy being correct, reliable,
and repeatable.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
At times, this conflicts with the need to innovate;
resolving it requires balancing trade-offs.

% FIXME: this should be expanded
Over the last few years there has been a massive increase in the use
of scientific Python due in part to the rapid growth of data science,
machine learning, and artificial intelligence.  
As a result, NumPy and its API has become ubiquitous.

@@

% Ralf: we should describe our role at the foundation of the stack
% i was thinking something along these lines ():
NumPy provides in-memory, multidimensional, homogeneously typed
(single pointer + strided) arrays on CPUs.
It supports many types of CPUs (e.g., POWER and ARM) with similar
performance as native C code.
While NumPy suits most user's needs, it cannot support every specialized need of the community;
due to its in-memory data model, it is limited in its ability to work with very
large datasets, datasets split across multiple systems, and computations that
require graphics processing units (GPUs), tensor processing units (TPUs) or
other specialized hardware for performance reasons.

% Ralf: should mention that there are typically utilities to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill the resultant gaps led to a
proliferation of array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

Importantly, some projects chose to follow the NumPy API exactly---with Dask,
CuPy, Sparse and JAX the most prominent examples---while others have
NumPy-inspired semantics but chose to deviate in places.
% Ralf: may want to add something like:
% when they deviate, they sometimes fix it, e.g.,
% https://github.com/pytorch/pytorch/pull/1983
% https://github.com/pytorch/pytorch/issues/2228
% many have for numpy users
% https://github.com/torch/torch7/wiki/Torch-for-Numpy-users

% \footnote{\url{https://cupy.chainer.org}}

% computation graphs---which postpone and combine calculations
% before executing them efficiently,

% Ralf: I am not sure where this goes, but it would be good to have something
% about dtype limitation in this section and i also think there should
% be some discussion of missing values somewhere
% (it is mentioned below, but not discussed)
While it supports a wide range of datatypes, it does not have good support
for missing values. % may be worth saying a few sentences about missing values
% if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)

@@

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402
To facilitate \emph{interoperability}, the NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.
